# Problem Set 1

## Measles 

(@) Load the __dslabs__ package and figure out what is in the `us_contagious_diseases` dataset. Create a data frame, call it `avg`, that has a column for `year`, and a `rate` column containing the cases of Measles per 10,000 people per year in the US. Because we start in 1928, exclude Alaska and Hawaii. Make sure to take into account the number of weeks reporting each year. If a week was not report, it should not be included in the calculation of the rate. 
```{r}
library(dslabs)
library(tidyverse)

# us_contagious_diseases %>% View

avg <- us_contagious_diseases %>% 
  filter(weeks_reporting > 0 & disease == "Measles" & !state %in% c("Alaska","Hawaii")) %>% 
  group_by(year) %>%
  summarize(rate = sum(count*52/weeks_reporting, na.rm = TRUE)/sum(population) * 10000)
```

(@) Use the data frame `avg` to make a trend plot showing the cases rate for Measles per year. Add a vertical line showing the year the Measles vaccines was introduced. Write a short paragraph describing the graph to someone you are urging to take the Measles vaccines.
```{r}
ggplot(data = avg) +
  geom_line(aes(x=year, y=rate)) + 
  geom_vline(xintercept = 1963, color = "red", linetype = "dashed") + 
  annotate("text", x = 1963, y = 60, label = "Measles Vaccine", hjust = -.1, color = 'red') +
  labs(title = "Measles rate per year in the United States", x = "Year", y = "Measles cases per 10,000 people")
```
The graph above is a trend plot of the rate of measles cases in the United States each year. The vertical line in red marks the introduction of the measles vaccine, after which there is a marked decline in the cases rate. The steady decline is followed by a plateau after the 1980s wherein the cases rate of measles is close to 0 cases/10,000 people. This trend suggests that the vaccine has contributed to decreasing the cases rate and preventing the spread of measles, and as such it is highly recommended to receive a measles vaccine.

(@) Is the pattern observed above the same for each state? Add a grey trend line for each state to the plot above. Use a transformation that keeps the high rates from dominating the figure.
```{r}
us_contagious_diseases %>% 
  filter(weeks_reporting > 0 & disease == "Measles") %>% 
  mutate(rate = count*52/weeks_reporting/population * 10000) %>% 
  ggplot() +
  geom_line(aes(group = state, x=year, y=rate), color = "grey", alpha = 0.5, na.rm = TRUE) +
  geom_line(aes(x=year, y=rate), data = avg) + 
  geom_vline(xintercept = 1963, color = "red", linetype = "dashed") + 
  annotate("text", x = 1963, y = 250, label = "Measles Vaccine", hjust = -.1, color = 'red') +
  scale_y_continuous(trans = "sqrt") +
  labs(title = "Measles rate per year in the United States", x = "Year", y = "Measles cases per 10,000 people")
```
No, the pattern observed above is not exactly the same for each state. Each state has its own trends of measles cases, but in general they do follow the same trend of a marked decline in cases after the introduction of the measles vaccine.

(@) In the plot above we can't tell which state is which curve. Using color would be challenging as it is hard if not impossible to find 48 colors we humans can distinguish. To make a plot where you can compare states knowing which is which, use one of the axis for state and the other for year, and then use hue or intensity as a visual cue for rates. Use a `sqrt` transformation to avoid the higher rates taking up all the color scale. Use grey to denote missing data. Order the states based on their highest peak.
```{r, fig.height=7}
library(RColorBrewer)
mycolors <- brewer.pal(9, "Blues")

us_contagious_diseases %>% 
  filter(!state %in% c("Hawaii","Alaska") & disease == "Measles") %>% 
  mutate(rate = count/population*10000*52/weeks_reporting,
         state = reorder(state, rate, max, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = state, fill = rate)) +  
  geom_tile(color = "grey") + 
  scale_x_continuous(expand = c(0,0)) +
  scale_fill_gradientn(colors = mycolors, trans = "sqrt") +
  geom_vline(xintercept = 1963, color = "red", linetype = "dashed") +  
  theme(panel.grid = element_blank(), legend.position = "bottom", text = element_text(size = 8)) +
  labs(title = "Measles rate per year per state in the U.S.", x = "Year", y = "State", fill = "Rate")
```

## COVID-19

(@) The csv file shared [here](https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-data.csv) includes weekly data on SARS-CoV-2 reported cases, tests, COVID-19 hospitalizations and deaths, and vaccination rates by state. 
 * Import the file into R without making a copy on your computer. 
 * Examine the dataset.
 * Write a sentence describing each variable in the dataset.
```{r}
library(readr)
url <- "https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-data.csv"
dat <- read_csv(url)
# str(dat)
```
There are 15 variables in the weekly COVID-19 dataset. State is a character variable which contains the same information as state_name, but in abbreviated form. Population is a numeric variable containing the population size of the state. Region is a numeric variable describing the geographic region to which a state belongs. mmwr_year and mmwr_week are numeric variables describing the epidemiologic year and week reported (CDC measures). Cases is a numeric variable describing the weekly number of cases. Tests is a cumulative numeric variable of the number of tests. hosp, deaths_underlying_cause, deaths_multiple_causes, and deaths_prov are also numeric variables. series_complete, booster, and bivalent are cumulative numeric variables, similar to tests, describing counts for vaccinated individuals.

(@) One of these columns could benefit from being a factor. Identify the column and convert it to factor.
```{r}
dat <- mutate(dat, region = factor(region))
# Could also do like this:
# dat$region <- factor(dat$region)
```

(@) Rather than providing a date, the dataset provides MMWR year and week. Read [this document](https://ndc.services.cdc.gov/wp-content/uploads/MMWR_Week_overview.pdf) and write a function to convert these to the start of the MMWR week in ISO-8601.
```{r}
library(lubridate)

mmwr_to_date <- function(mmwr_year, mmwr_week) {
  first_day <- floor_date(make_date(mmwr_year, 1, 4) , unit = "week")
  date <- first_day + weeks(mmwr_week - 1) 
  return(date)
}
```

(@) Add a columns `start_date` and `end_date` with the start and end of the MMWR week. Confirm that it worked by computing the MMWR week and year for both start and end date and comparing it to the MMWR week and year provided.
```{r}
dat <- dat %>% 
  mutate(start_date = mmwr_to_date(mmwr_year, mmwr_week),
         end_date = start_date + days(6))

# Confirm the following are TRUE
dat %>% summarize(w1 = all(epiweek(start_date) == mmwr_week),
                 y1 = all(epiyear(start_date) == mmwr_year),
                 w2 = all(epiweek(end_date) == mmwr_week),
                 y2 = all(epiyear(end_date) == mmwr_year))
```

(@) Make a trend plot similar to the one we made for Measles:

* Include a trend line for the US cases rate. Use per 100,000 person per week as the unit.
* Add a trend for each state, with color representing region.
* Use the end of the week date for your x-axis.
* Add a vertical dashed line on the day COVID-19 vaccination started being delivered in the US.

Write a paragraph describing the COVID-19 pandemic by describing the plot.
```{r}
covid <- dat %>% 
  group_by(end_date) %>% 
  summarize(rate = sum(cases, na.rm = TRUE)/sum(population) * 100000, .groups = 'drop')

dat %>% 
  mutate(rate = cases/population * 100000) %>%
  ggplot() +
    geom_line(aes(group = state, x=end_date, y=rate, color = region), na.rm = TRUE) +
    geom_line(aes(x=end_date, y=rate), data = covid, na.rm = TRUE) + 
    geom_vline(xintercept = as.Date('2020-12-14'), color = "red", linetype = "dashed") + 
    annotate("text", x = as.Date('2020-12-14'), y = 2500, label = "COVID-19 Vaccine", hjust = -.1, color = 'red') +
    labs(title = "COVID-19 cases rate per week in the United States", x = "Date", y = "COVID-19 cases per 100,000 people", color = "Region") 
```
The COVID-19 pandemic, similarly to measles, did not have the same exact same trend in case rate across all states. However, overall there are similar trends between the overall US cases rate and state rates, with a relative peak in cases in late 2020/early 2021 and an all-time peak in early 2022. In 2023, there appears to be a continual decline in the US cases rate.

(@) The number of cases depends on testing capacity. Note that during the start of the pandemic, when we know many people died, there are not that many cases reported. Also notice somewhat large variability across states that might not be consistent with actual prevalence. The `tests` columns provides the cumulative number of tests performed by the data represented by the week. This data is not official CDC data. It was provided by Johns Hopkins Coronavirus Resource Center. Before using the data, explore the data for each state carefully looking for potential problems.

For each state compute and plot the number of tests performed each week. Look at the plot for each state and look for possible problems. No need to make this plot pretty since we are just using it for data exploration. Report any inconsistencies if any.
```{r}
# Approach 1: Faceted plot with each state
dat %>%  
  filter(!is.na(tests)) %>% 
  arrange(end_date) %>% 
  group_by(state) %>% 
  mutate(tests = diff(c(0,tests))) %>% 
  ggplot(aes(end_date, tests/10^5)) + 
    geom_line() +
    facet_wrap(~state, scales = "free_y") + 
    labs(title = "COVID-19 tests per week per state", x = "Date", y = "COVID-19 tests per 100,000 people") 

# Approach 2: One plot with every state
dat %>% 
  filter(!is.na(tests)) %>% 
  arrange(end_date) %>%
  group_by(state) %>%
  mutate(tests = diff(c(0,tests))) %>%
  ungroup() %>% 
  ggplot() +
    geom_line(aes(group = state, x = end_date, y = tests/10^5), na.rm = TRUE) +
    labs(title = "COVID-19 cases rate per week in the United States", x = "Date", y = "COVID-19 cases per 100,000 people", color = "Region") 
```
The main inconsistency in the number of weekly tests is that in some cases we are seeing a negative number of tests because despite the 'tests' column being a cumulative measure, there are a few instances in which the number of tests decreases from one mmwr_week to the next within a state.

(@) To see if the inconsistencies seen in the previous plot are a problem, we are only going to look at the total number of tests at the end, plot the cumulative tests for each of the states with inconsistencies and see if the results are sensible. Explain your answer in 1-2 sentences.
```{r}
dat %>%  
  filter(!is.na(tests)) %>% 
  arrange(end_date) %>% 
  group_by(state) %>% 
  mutate(weeklytests = diff(c(0,tests))) %>%
  filter(any(weeklytests < 0)) %>% 
  ggplot() +
    geom_line(aes(group = state, x = end_date, y = tests/10^5, color = state), na.rm = TRUE) +
    labs(title = "Cumulative tests for states with inconsistent cumulative measures", x = "Date", y = "Number of tests per 100,000 people", color = "State")
```
The seven states with inconsistencies each have a dip in the cumulative tests measure at some point in the surveillance period, which aligns with the conclusions that we arrived to in our exploratory analyses.

(@) JHU stopped reporting testing some time in 2020. What was that date? Show the day in the format September 18, 2022.
```{r}
dat %>% 
  filter(!is.na(tests)) %>%
  summarize(`Last Day` = max(end_date)) %>%
  mutate(`Last Day` = format(`Last Day`, "%B %d, %Y")) 
```
(@) Compute the number of tests per capita for the last day JHU reported these statistics. Make a boxplot of these values for each region and include the state level data with the state abbreviation as a label. Write a sentences few describing these differences you see and how this could affect our interpretation of differences in cases rates across states. 
```{r}
last_day <- as_date("2021-12-18") 
dat %>% 
  filter(end_date == last_day) %>% 
  ggplot(aes(x = region, y = tests/population)) +
    geom_boxplot() +
    geom_text(aes(label = state), position = position_dodge(width = 0.8), vjust = -0.5) +
    labs(x = "Region", y = "Test per capita", title = "SARS-COV2 tests per capita in each U.S. region on last reported JHU day")
```
There are identifiable patterns in which these seems to be a systemically lower level of testing in certain regions of the United States. Overall, we see the largest number of tests per capita in Region 1 and the lowest in Region 7. For the regions in which we see more tests per capita, this must be considered as possibly influencing the cases rate to appear higher compared to a state which may have a lower rate in part because of less tests being administered.

(@) Although JHU stopped collecting testing data from the states, the CDC collected data from a few laboratories. We provide these date in [this url](https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-tests.txt).

* Import the data into R without downloading the file.
* Make sure that you create a data frame with a column with dates in `Dates` format and tests as numbers.
```{r}
url <- "https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-tests.txt"
tests <- read_delim(url, delim = " ") 
tests <- tests %>%  
  mutate(date = mdy(date), tests = parse_number(tests))
```

(@) Plot the tests per week to see the trend in time. Write a sentence of the pattern you see.
```{r}
ggplot(data = tests) +
  geom_line(aes(x = date, y = tests/10^4)) +
  labs(title = "COVID-19 tests per week", x = "Date", y = "Number of tests per 10,000 people")
```
The number of tests began to increase throughout the beginning of the pandemic before dropping again throughout 2021, increasing in the beginning of 2022, and we are now seeing a relatively steady drop since the beginning of 2023.

(@) The analysis on tests points to cases as not being a good measure of the state of the pandemic. Remake the trend plot but using death rates instead of cases rates. Write a sentence on how this plot better shows about the start of the pandemic that the cases plot did not. 
```{r}
death <- dat %>% 
  group_by(end_date) %>% 
  summarize(rate = sum(deaths_prov, na.rm = TRUE)/sum(population) * 100000, .groups = 'drop')

dat %>% 
  mutate(rate = deaths_prov/population * 100000) %>%
  ggplot() +
    geom_line(aes(group = state, x=end_date, y=rate, color = region), na.rm = TRUE) +
    geom_line(aes(x=end_date, y=rate), data = death, na.rm = TRUE) + 
    geom_vline(xintercept = as.Date('2020-12-14'), color = "red", linetype = "dashed") + 
    #annotate("text", x = as.Date('2020-12-14'), y = 2500, label = "COVID-19 Vaccine", hjust = -.1, color = 'red') +
    labs(title = "COVID-19 death rate per week in the United States", x = "Date", y = "COVID-19 deaths per 100,000 people", color = "Region") 
```
The trend plot of death rates reveals that deaths peaked at the beginning of the pandemic and were particularly high in the northeast regions, which is a distinct trend from our original cases rate plot that showed a peak in 'severity' during 2022.

(@) We want to examine the percent of the population that completed the first series of vaccines, received the booster, and received the bivalent booster, respectively. First run this line of code and examine what it does.

```{r}
tmp <- dat %>% 
  pivot_longer(c(series_complete, booster, bivalent), names_to = "series", values_to = "percent") %>% 
  select(state, region, population, end_date, series, percent) %>% 
  filter(!is.na(percent)) %>% 
  mutate(percent = percent/population,
         series = factor(series, c("series_complete", "booster", "bivalent"))) 
```
This line of code is pivoting the vaccine status types from columns to rows and generating a percent vaccinated out of the population per week for each state.

Then make a plot showing the percent of population vaccination for each state. Use color to represent region. 

* Show the dates on the x axis with the month abbreviation and year. 
* Place the three plots vertically, on top of each other. 
* Show percentages on the y axis. Hint: use `scales::percent`.
```{r}
ggplot(data = tmp) +
  geom_line(aes(group = state, x = end_date, y = percent, color = region)) +
  labs(title = "Percentage of population vaccinated per state", x = "Date", y = "Percentage vaccinated", color = "Region") +
  scale_x_date(labels = scales::date_format("%b %Y")) +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  facet_wrap(~series, ncol = 1)
```
(@) For each series used in the previous exercise, make a boxplot showing the maximum percentage reached by every state stratified by region. Let the range of the y axis adapt to the range of each series.
```{r, fig.height=7}
tmp %>% 
  group_by(state, series, region) %>% 
  summarize(max_percent = max(percent), .groups = 'drop') %>%
  ggplot(aes(x = region, y = max_percent)) +
    geom_boxplot() +
    # geom_text(aes(label = state)) +
    facet_wrap(~series, ncol = 1, scales = "free_y") +
    labs(x = "Region", y = "Maximum percentage vaccinated", title = "Maximum percentage vaccinated reached by each state")
```

(@) Use the plot above to define four periods: No vaccine, First vaccine, Booster, and Bivalent. 
Get the dates when each vaccine series starts (these don't have to be exact). Create a version of your data frame with a column called period that keeps the period associated with the week contained in each row.
```{r}
# Approximations:
# No vaccine: Jan 2020 
# First vaccine: Jan 2021 
# Booster: Oct 2021 
# Bivalent: Sept 2022 

period <- dat %>%
  mutate(period = if_else(end_date < as.Date('2021-01-01'), 'No vaccine',
            if_else(end_date < as.Date('2021-10-01'), 'First vaccine',
            if_else(end_date < as.Date('2022-10-01'), 'Booster', 'Bivalent'))))
```

(@) Define a new variable that stores the maximum vaccination percentage reached during each period. But for the first vaccine period use `series_complete`, for the booster period use the `booster` column, and for the bivalent period use the bivalent percentages. Remove data from the no vaccine period. Then make a plot comparing the COVID-19 death rate to the percent vaccinated. Use color to denote region. You should produce three plots, each with it's own axes range. Put the three plots in three entries of a 2x2 layout. Comment on what you observe.
```{r}
# Approach 1: Based on maximum end_date within each state-period grouping (cross-sectional at end of period)
period %>% 
  group_by(state) %>% 
  mutate(rate = deaths_prov/population * 100000) %>% 
  ungroup() %>%
  pivot_longer(c(series_complete, booster, bivalent), names_to = "series", values_to = "percent") %>% 
  select(state, region, population, end_date, rate, period, series, percent) %>% 
  filter(!is.na(percent)) %>% 
  mutate(percent = percent/population,
         series = factor(series, c("series_complete", "booster", "bivalent"))) %>%
  filter(period != 'No vaccine') %>%  # Remove data from the 'No vaccine' period
  group_by(state, period, series) %>%
  mutate(max_vaccination = max(percent)) %>% # Max vaccination percentage
  filter(
    (period == "First vaccine" & series == "series_complete") |
    (period == "Bivalent" & series == "bivalent") |
    (period == "Booster" & series == "booster")) %>% # Filter only for max percentages calculated according to instructions
  ungroup() %>%
  group_by(state, period) %>%
  slice(which.max(end_date)) %>% # Last date of period
  ggplot(aes(x = max_vaccination, y = rate, color = region)) +
  geom_point(na.rm = TRUE) +
  facet_wrap(~period, ncol = 2, scales = "free_y") +
  scale_x_continuous(labels = scales::percent_format(scale = 100)) +
  labs(title = "COVID-19 death rate vs vaccination percentage per state", y = "Death rate per 100,000 people", x = "Maximum vaccination percentage", color = "Region")

# Approach 2: Based on maximum death rate observed within each state-period grouping
period %>% 
  group_by(state) %>% 
  mutate(rate = deaths_prov/population * 100000) %>% 
  ungroup() %>%
  pivot_longer(c(series_complete, booster, bivalent), names_to = "series", values_to = "percent") %>% 
  select(state, region, population, end_date, rate, period, series, percent) %>% 
  filter(!is.na(percent)) %>% 
  mutate(percent = percent/population,
         series = factor(series, c("series_complete", "booster", "bivalent"))) %>%
  filter(period != 'No vaccine') %>%  # Remove data from the 'No vaccine' period
  group_by(state, period, series) %>%
  mutate(max_vaccination = max(percent)) %>% # Max vaccination percentage
  filter(
    (period == "First vaccine" & series == "series_complete") |
    (period == "Bivalent" & series == "bivalent") |
    (period == "Booster" & series == "booster")) %>% # Filter only for max percentages calculated according to instructions
  ungroup() %>%
  group_by(state, period) %>%
  slice(which.max(rate)) %>% # Max death rate
  ggplot(aes(x = max_vaccination, y = rate, color = region)) +
  geom_point() +
  facet_wrap(~period, ncol = 2, scales = "free_y") +
  scale_x_continuous(labels = scales::percent_format(scale = 100)) +
  labs(title = "COVID-19 death rate vs vaccination percentage per state", y = "Death rate per 100,000 people", x = "Maximum vaccination percentage", color = "Region")
```
In general, our plots support the logic we would expect wherein the COVID-19 death rate decreases as the maximum vaccination percentage increases.

(@) Does population density have an effect on infections? Use the `state.area` predefined variable to add compute population density. Make a histogram and density plot of state densities. Note that you will have to add the geographical area for Puerto Rico and DC as it is not included in `state.area`
```{r}
my.state.abb <- c(state.abb, "PR", "DC")
my.state.area <- c(state.area, 5325, 69)

popdens <- dat %>%  
  filter(end_date == min(end_date)) %>% 
  select(state, population) %>% 
  mutate(area = my.state.area[match(state, my.state.abb)]) %>%  
  mutate(popdens = population / area) 

popdens %>% ggplot(aes(popdens)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25, color = "black") +
  geom_density() +
  scale_x_log10() +
  labs(title = "Distribution of population density across states", x = "Population density", y = "Density")
```

(@) Plot death rates versus density for the four periods defined above.
```{r}
popdens <- popdens %>% select(-c(population))

# Approach 1: Based on maximum end_date within each state-period grouping (cross-sectional at end of period)
period %>% 
  left_join(popdens, by = 'state') %>%
  group_by(state, period) %>%
  mutate(rate = deaths_prov/population * 100000) %>% # death rate
  slice(which.max(end_date)) %>%
  ggplot() +
  geom_point(aes(group = state, x = popdens, y = rate, color = region), na.rm = TRUE) +
  facet_wrap(~period, scales = 'free_y') +
  scale_x_log10() +
  labs(title = "COVID-19 state population density vs. death rates", y = "Death rate per 100,000 people", x = "Population density", color = "Region")

# Approach 2: Based on maximum death rate observed within each state-period grouping
period %>% 
  left_join(popdens, by = 'state') %>%
  group_by(state, period) %>%
  mutate(rate = deaths_prov/population * 100000) %>% # death rate
  slice(which.max(rate)) %>%
  ggplot() +
  geom_point(aes(group = state, x = popdens, y = rate, color = region)) +
  facet_wrap(~period, scales = 'free_y') +
  scale_x_log10() +
  labs(title = "COVID-19 state population density vs. death rates", y = "Death rate per 100,000 people", x = "Population density", color = "Region")
```

